---
title: "Data task"
author: "Camilo Pedraza Jimenez"
date: "`r format(Sys.time(), '%B %d, %Y | %H:%M:%S | %Z')`"
output: 
  pdf_document:
    pandoc_args: --listings
    includes:
      in_header: preamble.tex
header-includes:
  - \usepackage{comment}
params:
  soln: TRUE
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = T}
#Load used packages
library(pacman)
pacman::p_load(tidyverse, ggplot2, vtable, rempsyc, knitr)
```

## Task 1

Load the Lalonde data set (see attachment) into either Stata or R. The
data belongs to a study that looked at the effectiveness of a job
training program (the treatment) on the real earnings of an individual,
a couple years after completion of the program. It consists of a number
of demographic variables as well as a treatment indicator and the real
earnings in the year 1978 (the response).

```{r}
lalonde  <- read.csv("lalonde.csv")
```

## Task 2

Produce a table with summary statistics (incl. mean, standard deviation)
for the variables in the dataset.

```{r, echo=TRUE, , fig.pos='H'}
lalonde %>%
  select(!nr) %>% #The variable nr includes only the number of the row, so it is not necessary for the summary statistics
  st(., simple.kable = T) #Function st() from the package vtable. simple.kable = T option to export the table to PDF format
```

Taking the first 5 variables as examples for the 445 observations in the dataset, we can see that: 
\begin{itemize}
  \item \textbf{Age:} The average age is 25 with a standard deviation of 7.1 years. 
  \item \textbf{Education:} The average years of education is 10 with a standard deviation of 1.8 years.
  \item \textbf{Black:} As a dummy variable, since the mean is above 0.5 (0.83), it is possible to say that the majority of people on the dataset are black. 
  \item \textbf{Hispanic:} As a dummy variable, since the mean is bellow 0.5 (0.088), it is possible to say that the majority of people on the dataset are not Hispanic. 
  \item \textbf{Married:} As a dummy variable, since the mean is below 0.5 (0.17), it is possible to say that the majority of people on the dataset are not Married. 
\end{itemize}
## Task 3

Produce two histograms showing the distribution of real earnings 1978
(re78) for individuals in the treatment and in the control group
separately.

```{r, echo=TRUE, fig.pos='H'}
ggplot(lalonde, aes(x = re78, fill = factor(treat))) +                      #Factor by treatment to get both histograms in one graph
  geom_histogram(position = "identity", alpha = 0.8, bins = 30) +           #alpha option to ajust the transparency
  scale_fill_manual(values = c("darkred", "grey"),
                    breaks = c(0, 1),
                    labels = c("Control", "Treatment")) +
  labs(title = "Distribution of real earnings in 1978 based on treatment",
       x = "Earnings",
       y = "Number of observations",
       fill = "Treatment type") +
  theme_minimal()
```

The distribution of the real earnings in 1978 is fairly even between the
treated and the control group, since the distribution of the earnings
for both groups stays mostly between 0 and 20000. Apart from the
similarity of both distributions, it is possible to notice that the
control group has more observation than the treatment group, since the
higher frequencies can be found in the control group. It is possible to
confirm that the control group is bigger by looking at the summary
statistics, in which the mean of the `treat` variable is 0.42, which is
lower than 0.5. Since the `treat` variable only takes the values 0 and
1, a mean lower than 0.5 would indicate that more control observations
are found in the dataset. Finally, the highest earnings can be seen
mostly for the treatment group, over 20000. The highest earnings for the
control (40000) and the treatment (around 35000 to 60000) groups.

## Task 4

Test whether the difference between treatment and control group is
statistically significant at conventional levels. What do you find?

```{r, echo=TRUE, fig.pos='H'}
nice_t_test(lalonde,              #nice_t_test() function from the rempsyc package for a summary of the t-test   
            response = "re78",
            group = "treat",
            warning = FALSE) %>%
  knitr::kable()                 #kable() function from the knitr package to get an organized table 
```

Based on the p-value (0.007893), it is possible to argue that, with a
level of confidence of 99%, we can reject the null hypothesis
($\mu_1 = \mu_2$ the mean of earnings for the control and treatment
group are equal). This means that we have sufficient evidence to say
that the mean of earning in 1978 between the two populations is
different. Although with the histogram graphs, the distribution of
earnings seemed similar, there could be two possible reasons that make
the difference in the average earnings significant:
\begin{itemize}
  \item The highest earnings of the treatment group.
  \item The high frequency of observations from the control group that have
earnings around 0.
\end{itemize}
